Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-04-05 manuscript/manuscript.tex
New section in manuscript: Methodology

\subsection{Analytical Solutions for Spherical Shell}

@@

Instead of increasing the GLQ orders to improve the accuracy, what leads to a more time consuming computation, adaptive discretization algorithms have been applied in previous works \citep{Li2011, Uieda2016}.
They essentially consist in dividing the tesseroid based on a ratio between the distance from its geometric centre to the computation point and its dimensions.

The main differences between the modified version made by \citet{Uieda2016} and the original one developed by \citet{Li2011} can be summed up in:
(1) a faster calculation of the distance from the computation point to the tesseroid and 
(2) the application of a stack based algorithm that speeds up the computation and gives more control over the recursion step.
More specifically, \citet{Uieda2016} redefined the dimensions of an arbitrary tesseroid: $L_\lambda$, $L_\phi$ and $L_r$. The former ones are the arc-distances measured along its top surface while $L_r$ is the difference between the radii of the top and bottom surfaces.

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance between the geometric centre of the tesseroid and the computation point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least one spherical coordinate, then the tesseroid must be divided by half in that direction.
A complete description of the modified adaptive discretization algorithm can be found in \citet{Uieda2016}.
On the contrary, if the inequality is satisfied for every spherical coordinate, the algorithm computes the desired gravity field using the GLQ approximation.

We will perform similar tests, but now with variable density spherical shell.
From the infinite density functions that can be proposed, we have chosen three characteristic ones: a linear, an exponential and a discontinuous function.

Specifically, the algorithm consists in the following steps:

\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
    \item Normalise the density function to the [0, 1] interval,
    \item Compute the absolute difference between the normalised density and a linear function that assumes the same values on the top and bottom radii that this normalised density.
    \item If the maximum of the absolute difference is greater that a predefined delta ratio ($\delta$), then the tesseroid will be subdivided on the radius at which this maximum takes place. Otherwise, the tesseroid won't be split.
\end{enumerate}

First we need to normalise the density function such as it assumes the values of 0 and 1 on the boundary radii ($R_1$, $R_2$), depending at which radius the original density function is maximum.
We call this normalised density $\rho_n(r')$ and we calculate it as follows:

\begin{equation}
    \rho_n(r') = \frac{\rho(r') - \rho_1}{\rho_2 - \rho_1},
\end{equation}
\noindent where
\begin{equation}
    \rho_1 = \text{min}\{ \rho(R_1), \rho(R_2) \}, \quad
    \rho_2 = \text{max}\{ \rho(R_1), \rho(R_2) \}.
\end{equation}

Secondly, we must define a linear function $\rho_l(r')$ that assumes the same values as the normalised density $\rho_n(r')$ on the boundary radii:

\begin{equation}
    \rho_l(r') =
    \frac{ \rho_n(R_2) - \rho_n(R_1) }{ R_2 - R_1 } (r' - R_1) + \rho_n(R_1),
\end{equation}

\begin{equation}
    \Delta \rho (r') = | \rho_n(r') - \rho_l(r') |.
\end{equation}

Finally, if the maximum of the absolute difference if greater than a predefined delta ratio $\delta$ we will divide the tesseroid on the radius at which this maximum takes place ($R_\text{max}$):

\begin{equation}
    \text{max}\{ \Delta \rho(r') \} > \delta,
    \label{eq:delta-density}
\end{equation}
\begin{equation}
    \text{max}\{ \Delta \rho(r') \} = \Delta \rho(R_\text{max}),
\end{equation}

In case that the original density function assumes the same value at both boundary radii, the procedure is the same except that the constants $\rho_1$ and $\rho_2$ are redefined as follows:

\begin{equation}
    \rho_1 = \text{min}\{ \rho(r') \}, \quad
    \rho_2 = \text{max}\{ \rho(r') \}.
\end{equation}

If on the contrary the inequation \ref{eq:delta-density} is not held, then the tesseroid will not be subdivided.
The same happens if the density function is constant or if the tesseroid radial dimension is bellow the threshold of 1mm.
On the first case due to the fact that the density-based discretization algorithm is not needed, and on the second one because smaller tesseroids will not introduce meaningful increases in the accuracy of the computation.

This algorithm can be subsequently applied to every subdividing tesseroid, until the inequation \ref{eq:delta-density} is not held or the subdivisions are smaller than 1mm.
This two conditions guarantee that the algorithm is finite.
Once this conditions are met, every subdividing tesseroid is subjected to the modified adaptive discretization algorithm and the GLQ in order to compute the gravity field generated by it.

It's easy to see that the delta ratio indirectly controls how many times the tesseroids will be subdivided based on the variation of the density function.
Therefore it determines the accuracy involving the numerical error that the density variation introduces and also the computation time.
This raises the need to determine a maximum value of $\delta$ that ensures an acceptable accuracy while minimising the computation time.
We will perform this determination with the same methodology as the one applied to the $D$ ratio: compare the numerical model with the analytical solutions for a spherical shell.

>>>>

\section{Methodology}

From equation \ref{eq:glq-var-dens} we can conclude that the error of 
the numerical approximation is controlled by the order of the 
quadrature, i.e. the number of point masses that are going to be used 
to compute the gravitational field.
This conclusion could lead to a naive methodology: find the order of 
the quadrature that produces gravitational fields with an acceptable 
accuracy.
Although this simple proposal has a few problems: (a) it won't be 
efficient and (b) it couldn't be applied to the general case (any
tesseroid with arbitrary density function on any computational point) 
without modifying the quadrature order taking into account 
characteristics of the tesseroid and the computation point (tesseroid 
size, distance to the computation point, variation of the density 
function, etc).

That's why we will keep the GLQ order fixed and make use of 
discretization algorithms.
Instead of uniformly increase the number of point masses that 
approximates a certain tesseroid, the discretization algorithms are 
intended to efficiently subdivide the tesseroid into smaller ones and 
apply the GLQ to each one of them.
These algorithms generally involve a discretization inequation that 
controls if the tesseroid should be subdivided or not and a certain 
predefined parameter that works as a threshold for the discretization 
inequation.

On the variable density tesseroid problem we found two main error 
sources in the GLQ that should be addressed: (a) the relation between 
the tesseroid size and its distance to the computation point, and (b) 
the variation of the density function.

The former one is also present in the homogeneous density tesseroid 
models, such as the one developed by \citet{Uieda2016}, and could be 
solved by applying a modified version of the adaptive discretization 
algorithm proposed by \citet{Li2011}.
Although for the latter we need to introduce a new density-based 
discretization algorithm that subdivides the tesseroid taking into 
account the variation of its density function.

Studying the homogeneous density tesseroid problem, \citet{Ku1977} 
noticed that keeping the GLQ orders fixed, the approximation 
becomes less accurate when the computation point is closer to the 
tesseroid, making the point masses effect more noticeable than the 
tesseroid we want to approximate.
A complete analysis of this conclusion can be found in \citet{Ku1977, 
Li2011, Uieda2016}.

Instead of increasing the GLQ orders to improve the accuracy, what 
leads to a more time consuming computation, \citet{Li2011} proposed an 
adaptive discretization algorithm which essentially consists in 
dividing the tesseroid based on a ratio between the distance to the 
computation point and its dimensions.
\citet{Uieda2016} proposed a modified version of the algorithm proposed 
by \citet{Li2011}. These modifications can be summed up in:
(1) a faster calculation of the distance from the computation point to 
the tesseroid and 
(2) the application of a stack based algorithm that speeds up the 
computation and gives more control over the recursion step.
More specifically, \citet{Uieda2016} redefined the dimensions of an 
arbitrary tesseroid: $L_\lambda$, $L_\phi$ and $L_r$. The former ones 
are the arc-distances measured along its top surface while $L_r$ is the 
difference between the radii of the top and bottom surfaces.

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance 
between the geometric centre of the tesseroid and the computation 
point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least 
one spherical coordinate, then the tesseroid must be divided by half in 
that direction.
On the contrary, if the inequality is satisfied for every spherical 
coordinate, the algorithm computes the desired gravity field using the 
GLQ approximation.
A complete description of the modified adaptive discretization 
algorithm can be found in \citet{Uieda2016}.

We will perform similar tests, but now with variable density spherical 
shell.
From the infinite density functions that can be proposed, we have 
chosen three characteristic ones: a linear, an exponential and a 
discontinuous function.
\todo{sacaria todo este parrafo y lo introduciria mas adelante}

In a general way, the algorithm can be summarised in the following steps:

\begin{enumerate}
\renewcommand{\theenumi}{(\arabic{enumi})}
    \item Normalise the density function to the [0, 1] interval,
    \item Compute the absolute difference between the normalised 
          density and a linear function that assumes the same values on the 
          top and bottom radii that this normalised density.
    \item If product of the maximum of the absolute difference and a 
          tesseroid size ratio is greater that a predefined delta ratio 
          ($\delta$), then the tesseroid will be subdivided on the 
          radius at which this maximum takes place. Otherwise, the 
          tesseroid won't be split.
    \item If it's discretized, then apply items (2), (3) and (4) for 
          every subdividing tesseroid.
\end{enumerate}

Lets assume an arbitrary tesseroid with boundary radii $R_1$ and $R_2$ 
and a density function $\rho(r')$ which we will call the 
\emph{original tesseroid}, whose radial dimension can be written as
$L_r = R_2 - R_1$.

First we need to normalise the density function such as its minimum and 
maximum will be 0 and 1, respectively.
We call this normalised density $\rho_n(r')$ and we calculate it as 
follows:

\begin{equation}
    \rho_n(r') = \frac{\rho(r') - \rho_1}{\rho_2 - \rho_1},
\end{equation}
\noindent where
\begin{equation}
    \rho_1 = \text{min}\{ \rho(r') \}, \quad
    \rho_2 = \text{max}\{ \rho(r') \} \quad
    \forall \, r' \in [R_1, R_2].
\end{equation}

\noindent Because this algorithm will be iteratively applied for every 
subdividing tesseroid, we'd like to note that this normalised density 
function will be the same for every subdividing tesseroid.
In case that the density function is constant, both the maximum and 
minimum densities are equal and the normalised function could not 
be defined, so the density-based discretization algorithm won't be 
applied.

Secondly, we must define a linear function $\rho_l(r')$ that assumes 
the same values as the normalised density $\rho_n(r')$ on the boundary 
radii $R_1$ and $R_2$:

\begin{equation}
    \rho_l(r') =
    \frac{ \rho_n(R_2) - \rho_n(R_1) }{ R_2 - R_1 } (r' - R_1) + \rho_n(R_1),
    \label{eq:density-reference-line}
\end{equation}

\begin{equation}
    \Delta \rho (r') = | \rho_n(r') - \rho_l(r') |.
    \label{eq:density-abs-diff}
\end{equation}

Finally, if the product of the maximum of the absolute difference and a 
tesseroid size ratio is greater than a predefined delta ratio $\delta$ 
we will divide the tesseroid on the radius at which this maximum takes 
place ($R_\text{max}$):

\begin{equation}
    \text{max}\{ \Delta \rho(r') \} \frac{L_r^\text{sm}}{L_r} > \delta,
    \label{eq:delta-density}
\end{equation}
\begin{equation}
    \text{max}\{ \Delta \rho(r') \} = \Delta \rho(R_\text{max}),
\end{equation}

The $L_r^\text{sm}$ is the radial dimension of the tesseroid that is 
currently analysed. On the first iteration, this tesseroid is the 
\emph{original} one, so $L_r^\text{sm} = L_r$, but for future 
iterations this radial dimension will be computed as the difference of 
the top and bottom radii of the subdividing tesseroid.

This algorithm will be subsequently applied to every subdividing 
tesseroid.
We will start from equation \ref{eq:density-reference-line} defining a 
new linear function $\rho_l(r')$, an absolute difference with equation 
\ref{eq:density-abs-diff} and evaluate the discretization inequation 
\ref{eq:delta-density}, taking into account that now $R_1$ and $R_2$ 
are the radial boundaries of this smaller tesseroid.

The tesseroid size ratio included in the inequation 
\ref{eq:delta-density} is intended to reduce the number of probable 
unneeded discretizations and focus on discretizations that will 
significantly reduce the numerical error.
A big subdiving tesseroid with small maximum absolute difference would 
introduce more numerical error than a small subdividing tesseroid with 
higher maximum absolute difference.

If on the contrary the inequation \ref{eq:delta-density} is not held, 
then the tesseroid will not be subdivided.
The same happens if the tesseroid radial dimension is bellow the 
threshold of 1mm. On the first case due to the fact that no more 
discretizations are needed, and on the second one because smaller 
tesseroids will not introduce meaningful increases in the accuracy of 
the computation.
This two conditions guarantee that the algorithm is finite.
Once they are met, every subdividing tesseroid is subjected to the 
modified adaptive discretization algorithm and the GLQ in order to 
compute the gravity field generated by it.

It's easy to see that the delta ratio indirectly controls how many 
times the tesseroids will be subdivided based on the variation of the 
density function.

Therefore it determines the accuracy involving the numerical error that 
the density variation introduces and also the computation time.
This raises the need to determine a maximum value of $\delta$ that 
ensures an acceptable accuracy while minimising the computation time. 
We will perform this determination with the same methodology as the one 
applied to the $D$ ratio: compare the numerical model with the 
analytical solutions for a spherical shell.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Analytical Solutions for Spherical Shell}

@@
